---
execute: 
  echo: true
  eval: false
  warning: false
---

# Chapter title


```{r here}
#| message: false
#| eval: true

library(here)

set.seed(123) # for reproducibility
```

## Section 1


## Section 2


## Session info {.unnumbered}

This document was created under the following conditions:

```{r sessioninfo}
#| eval: true
#| echo: false
sessioninfo::session_info()
```

## References {.unnumbered}

# Masterproject : Structural Divergence 

# Packages required
This analysis uses the following libraries (including their respective versions): bio3d ver. 2.4.5, R.utils ver. 2.13.0, ggplot2 ver. 4.0.2, tidyverse ver. 2.0.0, clusterProfiler ver. 4.14.6, ggstatsplot ver. 0.13.5. 
```{r setup}
#| echo: false
# Load libraries
library(bio3d)
library(R.utils)
library(ggplot2)
library(tidyverse)
library(clusterProfiler)
library(ggstatsplot)
library(performance)
library(rstatix)
library(conover.test)
library(rcompanion)
library(report)
library(sessioninfo)
# Set path to working directory
setwd("C:/OneDrive/OneDrive - UGent/Academiejaren/jaar5_semester2/Masterproject/Data")
```

# File digestion

## Removing .cif files
Upon researching and downloading the whole proteome of an organism from the Alphafold server, the Alphafold server will provide the proteome in .cif and .pdb files. In order to save space on the hard drive of the computer, the .cif-files are first removed.
```{r}
# Set your directory path - Don't forget to replace organism to organism of interest!
dir_path <- "./Alphafold/Z.maize"

# List all .cif files
cif_files <- list.files(path = dir_path, 
                        pattern = "\\.cif.gz$", 
                        full.names = TRUE)

# Remove all .cif files 
file.remove(cif_files)
```
With the .cif-files removed, the remaining .pdb files are unpacked for further research.

## Unzipping .pdb files 
In order to optimally access the content of the .pdb files, the files are first unzipped. The gunzip has a setting that normally allows the zipped file to be removed upon unpacking. This can take a relatively long time. Alternatively, the unzipped files can be moved to a different file e.g. here called `unzipped`, after which the zipped files can be removed manually. This allows the operations to be split, reducing chances for the system to crash while computing. 
```{r}
# List all .gz files
gz_files <- list.files(path = dir_path,
                       pattern = "\\.gz$",
                       full.names = TRUE)

# unzipped_dir <- file.path(dir_path, "unzipped")
# if (!dir.exists(unzipped_dir)) dir.create(unzipped_dir)

# Loop and unpack
for (file in gz_files) {
  output_file <- file.path(
    dir_path,
    sub("\\.gz$", "", basename(file))
  )
  
  gunzip(file,
         destname = output_file,
         overwrite = TRUE,
         remove = TRUE)
}
```
# Structural comparison 
In this study, the structural comparison between a pair of proteins is based on a global RMSD-calculation, based on the C-alpha backbone of the proteins, through the use of the functions `pdbaln()` followed by `atom.select()` and `rmsd()`, all being from the `bio3d` package. The functions are enveloped in a larger function (`calculate_rmsd`) for high-throughput analysis of all protein-pairs. At the same time, this larger function will also retrieve the sequence similarity between the protein sequences within a pair, retrieved from the `pdbaln()` function, for visualizing the relation between the RMSD and sequence similarity later on. 

## Loading of the dataframe 
A dataframe is set up containing the information coming from the Doubletrouble-database. Before loading the information of this database however, the EnsembL ID's are converted into Uniprot ID's. This conversion is necessary to link to the proteins to the .pdb structures gathered from the Alphafold database. Three columns are added to the dataframe for storing the results coming from `calculate_rmsd` function (see further): namely, the sequence identity, RMSD, and the B-factor after structural analysis is added per pair of pdb's. The B-factor computation is applied to deduce wether there is confounding of the B-factor with the RMSD. In Alphafold, the B-factor column of the .pdb's is applied to express Alphafold's confidence in predicting the structure. The higher this result, the more confidence Alphafold has in its prediction.
```{r}
# Read file coming from Doubletrouble database, in which the entries are converted from Ensembl to Uniprot database conversion
data <- read.delim("./Doubletrouble/zma_pairs_uniprot.tsv", stringsAsFactors = FALSE) |>
  # During the conversion of Ensembl to Uniprot, several duplicate pair entries were duplicated.
  # Through the distinct-function, these duplicate pairs of duplicates are filtered out to retrieve, to consider only one of these duplicates. As such, each pair of duplicates is considered once, which is ideal for statistical analysis 
  distinct() |>
  # During the conversion of Ensembl to Uniprot, several duplicate pair entries had instances where dup1 would be equal to dup2. These are filtered out. 
  filter(dup1 != dup2)
# Create empty numeric columns for storing the sequence identity (in seq_id), rmsd between pdb's (in rmsd) and average B-factor
data$seq_id <- NA_real_
data$rmsd <- NA_real_
data$bavg <- NA_real_
```

## Computing RMSD- and sequence-similarity aswell as B-factor values for each pair of duplicates
The RMSD is computed using the `rmsd()` function from the `bio3d` library. This function requires a sequence alignment of the protein sequences, which can be performed by `pdbaln()`, coming from the same library. To perform the sequence alignment, the library requires an older version of `MUSCLE` (ver.	`muscle3.8.31`), which can be downloaded at https://drive5.com/muscle/downloads_v3.html. For further installation instructions, consult the "*installing Bio3D*" section at the following link: http://thegrantlab.org/bio3d_v2/tutorials/installing-bio3d.  From the `pdbaln()` function also allows extraction of the sequence similarity and the B‑factor values. Although the B‑factor is an inherent property of a .pdb file, the B‑factors extracted through`pdbaln()` correspond only to the atoms that were used during the structural alignment. Considering we are comparing the global fold between two homologues protein structures, it has been opted to select for Calpha-atoms only. This reduces inflation of the RMSD-value coming from differences in side-chain positioning between two structures. Therefore, not all atoms from the original structures are included. Additionally, if the two protein sequences differ in length, certain residues will not participate in the alignment, further reducing the set of B‑factors returned.
```{r}
# Specify function
get_values <- function(pdb1, pdb2){
# Specify the paths of both pdb's 
pdb1path <- paste0("C:/OneDrive/OneDrive - UGent/Academiejaren/jaar5_semester2/Masterproject/Data/Alphafold/Z.maize/AF-", pdb1, "-F1-model_v6.pdb")
pdb2path <- paste0("C:/OneDrive/OneDrive - UGent/Academiejaren/jaar5_semester2/Masterproject/Data/Alphafold/Z.maize/AF-", pdb2, "-F1-model_v6.pdb")
# Read PDB-file
# The ATOM.only = TRUE is useful for speed enhancements with large files where secondary structure, biological unit and other remark records are not required. This does not influence the output necessary for this experiment. 
pdb1 <- read.pdb(pdb1path, ATOM.only = TRUE)
pdb2 <- read.pdb(pdb2path, ATOM.only = TRUE)

#Select for C-alpha atoms 
ca1 <- atom.select(pdb1, "calpha")
ca2 <- atom.select(pdb2, "calpha")

# Perform alignment
aln <- pdbaln(list(pdb1, pdb2), fit = FALSE, atom.select = list(ca1, ca2))
# Retrieve sequence identity from alignment 
# The seqidentity-function returns a 2 x 2 matrix based on the alignment. One element is specified to be retrieved from this matrix that compares pdb1 vs. pdb2
id <- seqidentity(aln)[1, 2]

# pdbaln is a wrapper function that takes on multiple tasks. In order to relieve some of the computational load, the fitting will be performed within the RMSD-function. 
# If the fit-parameter within the rmsd-function is set to TRUE, the RMSD coordinate superposition is performed prior to RMSD calculation. As such, it is expected that the RMSD-function will take on to fit the sequences itself
# The RMSD-function returns a 2 x 2 matrix based on our alignment. One element is specified to be retrieved from this matrix that compares pdb1 vs. pdb2
rmsd_val <- rmsd(aln, fit=TRUE)[1, 2]

# Extract the B-factors per compared pair of atoms within the alignment 
# use apply() function across column to identify which column has two B-factor entries, meaning these were structurally aligned, from columns that contain only one B-factor entry, meaning the structures were not compared at that position
aligned_cols <- apply(aln$b, 2, function(col) all(!is.na(col)))
# Retrieve the B-factors of the positions that were structurally aligned and compared
aligned_scores <- aln$b[, aligned_cols]
# Get the average of the B-factors
bavg <- mean(aligned_scores)

# Return the computed RMSD- and sequence identity values in the form of a vector 
return(c(rmsd_val, id, bavg))
}
```

## Iterating through data
The sequence similarity, retrieved from the `pdbaln()`function, and the RMSD-value, from the `rmsd()` function, are computed for each pair of duplicates and stored within the dataframe through a `for`-loop. The former two functions are enveloped in a larger function called `calculate_rmsd()`. If `calculate_rmsd()` fails to perform any of the former functions (e.g. due to the abscence of a UniProt structure within the Alphafold database), NA-values are set to replace any errors. This allows the function to continue and iterate over all the rows within the dataframe, without any errors interrupting the `for`-loop. 
```{r}
# Iterate through each row
for (row in 1:nrow(data)) {
  # At row x retrieve the name of the pdb-file at resp. column "dup1" and "dup2"
  pdb1 <- data[row, "dup1_uniprot"]
  pdb2 <- data[row, "dup2_uniprot"]
  
  # tryCatch enables to continue the iteration even though errors might occur
  vals <- tryCatch(
    # Execute expression; in this case, the function calculate_rmsd
    expr = {get_values(pdb1, pdb2)},
    
    # If an error occurs, return 'NA'
    error = function(e){return(NA)}
  )
  # The values obtained from the function are stored within the dataframe 
  data$rmsd[row] <- vals[1]
  data$seq_id[row] <- vals[2]
  data$bavg[row] <- vals[3] 
}
```

# Visualization
With the data digested and values computed, this section aims to bring together the data into a concise visualization. 

## Data prepping
Before visualization, some tweaks still have to be made to the data in order to avoid getting warnings and to maintain consistency throughout the figures. For the former, the entries that failed to produce an RMSD-value, therefore containing an `NA`-value within the `rmsd`-column, are removed. As for the latter, the types of duplications are ordered, ensuring that the order throughout the figures remains the same. 
```{r}
data_ordered <- data |>
  # Remove homologues pairs that failed to compute an RMSD-value
  filter(!is.na(rmsd)) |>
  # Order the data based on type of duplication
  transform(type=factor(type,levels=c("SD","TD","PD", "rTRD", "dTRD", "DD")))
```

## Plotting RMSD vs. type of duplication
To gain better insight on how the RMSD varies between the types of duplications, the RMSD is plotted as a violin plot per type of duplication. Besides that, the differences between groups is also statistically evaluated. 

### Statistical analysis 
It is important to use an appropriate test, that has the most power, to which the data meets the required assumptions. Multiple transformations were attempted with the data, but repeatedly normality was not met. That is why the significance of the data will be investigated using a Kruskal-Wallis test. This test requires the presence of three or more groups within the data and expects the dependent variable to be at least on the ordinal scale. Considering the distributions differ in shape and variance (see figure p), the Kruskal-Wallis test will test differences in distributions of the RMSD between different groups.
```{r}
kruskal_test(rmsd ~ type, data = data)
```
The p-value is less than the significance level 0.05, therefore there are significant differences between the types of duplications in terms of the distribution of RMSD. 

### Post-hoc testing
Considering the Kruskal-Wallis test is significant, the analysis continues with post-hoc testing. For this a Conover-Iman test will be applied. The Conover-Iman gains preference over the Dunn's test for its increased power (without requiring additional assumptions), therefore being able to avoid increased probability of having a type II error. Given that the sample size of each group is at least n = 660, the dataset is expected to do well with this test. 
```{r}
conover <- conover.test(data$rmsd , data$type, method="holm")
```
Using Holm-Bonferroni to adjust the p-values, the Conover-Iman test found almost every comparison between two groups significant, except for the comparison between dTRD - rTRD and PD - TD. 

### Compact letter display
In order to provide a visual summary of the significance groupings across all pairwise comparisons, a compact letter display (CLD) is applied to assign letters that represent which groups are statistically indistinguishable. 
```{r}
# Generate a CLD using the cldList function from the rcompanion-library
cld <- cldList(comparison = conover$comparisons,
               p.value    = conover$P.adjusted,
               threshold  = 0.05) |>
  # Select relevant columns for the visual representation
  select(Group, Letter) |>
  # Capitalize the letter-notation
  mutate(Letter = toupper(Letter), pos_y = max(data_ordered$rmsd))
```

### Plot
Finally, with the compact letter display computed and the data rendered, the plot can be created that expresses RMSD vs. the type of duplication. 
```{r}
figure1 <- ggbetweenstats(data_ordered, x = type, y = rmsd,
               # Specify statistical testing to be nonparametric 
               type = "nonparametric",
               # Do not show the standard pairwise display
               pairwise.display = "none",
               # Do not annotate means 
               centrality.plotting = FALSE,
               # Hide results displayed on subtitle
               results.subtitle = FALSE,
               # Change titles for x- and y-axes respectively
               xlab = "Duplication type",
               ylab = "RMSD (Å)",
               # Manually arrange theme
               ggtheme = ggplot2::theme(
                 # Hide grid
                 panel.grid = element_blank(),
                 # Specify axes lines' color and size 
                 axis.line = element_line(color = "black", size=0.8),
                 # Specify background of panel and provide black outline
                 panel.background = element_rect(fill = 'white'),
                 # Change size on x- and y-axes respectively
                 axis.text.x = element_text(color = "black", size=12),
                 axis.text.y = element_text(color = "black", size=12)
               )) + 
  # Add an extra layer of geom_label to specify the CLD
  geom_label(
    # Specify CLD-dataframe computed in section '### Compact Letter Display'
    data  = cld,
    # Specify aesthetics
    aes(x = Group, y = pos_y, label = Letter),
    # Avoid inheriting the aesthetics set previously at ggbetweenstats 
    inherit.aes = FALSE,
    # Increase size
    size  = 8,
    # Set the CLD lower than the anchoring point in order to not block the outlier at DD
    vjust = 1.5
  )
```

## Plotting of RMSD vs. sequence identity based on type of duplication
To further visualize how the RMSD value varies throughout the duplicate pairs being compared, as well as the sequence identity, the RMSD is plotted against the sequence identity per type of duplication. 
```{r}
# In order to maintain consistency of colors of the types of duplications throughout the report and this file, the colors are stored to apply in future figures. This is done by extracting the buildup of figure p and retrieving the colors assigned to the groups of the data
cols <- unique(ggplot_build(p)$data[[1]]$colour)

# Specify RMSD grouped by duplication type, color based on duplication type
figure2 <- ggplot(data_ordered, aes(x = seq_id, y = -log10(rmsd))) +
  # Specify violin plot, change outline's color and size
  geom_point(size = 2, shape = 1, aes(color = type)) +
  # Specify the colors manually in order to maintain consistency throughout the figures 
  scale_colour_manual(values = cols) +
  # Apply a trendline throughout the plot in order to visualize the general behavior of the datapoints
  geom_smooth(se = TRUE, aes(group=type), alpha = 0.75, color = "black") +
  # Create a separate plot per type in order to improve visualization
  facet_wrap(vars(type)) +
  # Specify geom_point to have empty dots 
  scale_shape(solid = FALSE) +
  # Specify x- and y-titles 
  labs(x = "Sequence identity (%)", y = "RMSD (Å)") +
  # Customize theme
  theme_minimal() +
  theme(
    # Hide legend
    legend.position = "none",
    # Specify background of panel and provide black outline
    panel.background = element_rect(colour = 'black', size=0.8),
    # Remove grid
    panel.grid = element_blank(),
    # Change size on x- and y-axes respectively
    axis.text.x = element_text(color = "black", size=12),
    axis.text.y = element_text(color = "black", size=12),
    # Specify axes lines' color and size 
    axis.line = element_line(color = "black", size=0.8),
    # Add ticks along the labels on the x- and y-axes
    axis.ticks.x = element_line(color = "black", size=0.8),
    axis.ticks.y = element_line(color = "black", size=0.8),
    # Specify tick length
    axis.ticks.length = unit(0.15, "cm"),
    # Change size of plot subtitles 
    strip.text = element_text(size = 10))
figure2
```

# Functional enrichment analysis 
The functional enrichment analysis consists of the analysis of the top 10% most divergent proteins and the top 10% least divergent proteins (based on RMSD-calculation) and is aimed to evaluate enriched biological processes, Reactome pathways, and protein domain categories.

## Data preparation
The first step is to index the data based the sorted RMSD-value and remove entries where there's no RMSD-value `(rmsd = NA)`.
```{r}
# Index the data based on rmsd-value
data_idx <- data_ordered |>
  # Sort based on rmsd 
  arrange(rmsd) |>
  # Index based on sorting based on rmsd 
  mutate(idx = row_number())
```

## Administer STRING-ID's
The STRING database is applied to denote the biological process, reactome pathways and protein domains. The database makes use of STRING-ID's to streamline the ID's coming from different databases. The name of the protein there has to be linked to the protein STRING-id within the database. This conversion is mediated through the protein aliases dataframe, acquired from the STRING database.
```{r}
# Load STRING protein aliases
df_aliases <- read_tsv(
  "https://stringdb-downloads.org/download/protein.aliases.v12.0/4577.protein.aliases.v12.0.txt.gz",
  show_col_types = FALSE) |>
  # Subset aliases from STRING protein aliases file 
  subset(select = c("#string_protein_id", "alias")) 

# Rename columns 
colnames(df_aliases) <- c("string_id", "alias")

# Left_join() from the dplyr-library is applied to match the protein name in data with the STRING-id stored within the database. This is applied to both duplicates, starting with dup1
data_idx <- left_join(data_idx, df_aliases, by = c("dup1_uniprot" = "alias")) |> 
  # Left_join appears to duplicate the entries, therefore; apply distinct() to select for distinct rows
  distinct() |> 
  # Rename the recently added column 
  rename("string_id_dup1" = "string_id")

# The same is applied to dup2
data_idx <- left_join(data_idx, df_aliases, by = c("dup2_uniprot" = "alias")) |>
  distinct() |>
  rename("string_id_dup2" = "string_id")

# df_aliases is no longer required, in order to have a clean Workspace, it is advised to remove it
rm(df_aliases)
```

## Compute background and top 10%
As discussed earlier the the top 10% most divergent proteins and the top 10% least divergent proteins (based on RMSD-calculation) undergo functional enrichment to evaluate enriched biological processes, Reactome pathways, and protein domain categories. In this section, these entries are collected. Simultaneously, the background will be computed for the `enricher()` function of the `clusterProfiler` library. 
```{r}
# Select the top 10% rows at the end, or at the beginning, of the dataframe for functional enrichment analysis 
lower <- head(data_idx, n = round(nrow(data_idx) * 0.1))
upper <- tail(data_idx, n = round(nrow(data_idx) * 0.1))

# Pool all STRING ID's to form the background for the enricher() function for enrichment analysis
# Only select unique STRING ID's to avoid duplicates 
bg <- c(data_idx$string_id_dup1, data_idx$string_id_dup2) |> unique()
```

## Intermezzo: visualization of RMSD-value vs. index
In order to visualize the datapoints that will be considered during functional enrichment analysis, all indexed points are plotted against their respective RMSD-values. The top 10% lowest and highest scoring datapoints based on RMSD-value, will be marked on the plot to show the reader the data that is considered for the analysis. 
```{r}
data_idx <- data_idx %>%
  mutate(color_group = case_when(
    idx %in% lower$idx ~ "lower",
    idx %in% upper$idx ~ "upper",
    TRUE               ~ "other"   # everything else is black
  ))


figure3 <- ggplot(data_idx, aes(x = idx, y = rmsd, color = color_group, size = color_group)) +
  geom_point(alpha = 0.9) +
  scale_color_manual(
    values = c(
      lower = "#1E64C8",
      upper = "#EB3324",
      other = "black"
    )) +
  scale_size_manual(values = c(
    lower = 2.8,
    upper = 2.8,
    other = 2.0
  )) +
  # scale_size_manual(values = c(`FALSE` = 1.5, `TRUE` = 3)) +
  labs(
    x = "Index",
    y = "RMSD (Å)",
    color = "Covered",
    size = "Covered"
  ) +
    theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.text.x = element_text(color = "black", size=12),
    axis.line = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),

    axis.text.y = element_text(color = "black", size=12),
    axis.ticks.y = element_line(color = "black"),
    
    axis.ticks.length = unit(0.10, "cm")
  )
figure3
```

## Retrieval of enrichment terms
As mentioned previously, the functional enrichment analysis will be performed using the STRING database. To do this, a dataframe containing all enrichment terms for Zea mays is loaded into the R environment. From this dataframe, the columns of interest (that being, biological processes, reactome pathways and protein domains) are selected. 
```{r}
# Load in the STRING protein enrichment terms file of Zea mays into the R environment 
df_enrichment_terms <- read_tsv(
  "https://stringdb-downloads.org/download/protein.enrichment.terms.v12.0/4577.protein.enrichment.terms.v12.0.txt.gz", 
  # Avoid showing output
  show_col_types = FALSE) |>
  # Select the columns relevant for this analysis 
  subset(select = c("#string_protein_id", "category", "description"))

# Rename the subsetted dataframe's columns for ease-of-use
colnames(df_enrichment_terms) <- c("string_id", "category", "description")

# Create a dataframe that contains everything of the category: Biological Process
biologicalprocess <- df_enrichment_terms[
  df_enrichment_terms$category == "Biological Process (Gene Ontology)",] |>
  # Remove the 'category' column from the dataframe
  subset(select = -category)

# Create a dataframe that contains everything of the category: Reactome Pathways
reactomepathway <- df_enrichment_terms[
  df_enrichment_terms$category == "Reactome Pathways",] |>
  # Remove the 'category' column from the dataframe
  subset(select = -category)

# Create a dataframe that contains everything of the category: Protein Domains
proteindomain <- df_enrichment_terms[
  df_enrichment_terms$category == "Protein Domains (Pfam)",] |>
  # Remove the 'category' column from the dataframe
  subset(select = -category)

# Collect all seperate categories (Biological Process, Reactome Pathways and Protein Domains), into one dataframe 
# Paste an abbreviation next to each term to allow for easier identification and literature research
annot <- rbind(
 biologicalprocess |> mutate(description = paste0("bp:", description)), 
 reactomepathway |> mutate(description = paste0("rp:", description)), 
 proteindomain |> mutate(description = paste0("pd:", description))) |>
 select(2, 1)

rm(df_enrichment_terms)
```

## Enrichment analysis 
With the top 10% lowest and highest scoring entries based on RMSD-value specified in separate dataframes and the enrichment terms specified ready for analysis, functional enrichment can be performed. 
```{r}
# Pool all duplicates into one vector and select unique entries to avoid duplicates 
lower_id <- c(lower$string_id_dup1, lower$string_id_dup2) |> unique()

# Perform enrichment analysis on the top 10% lowest-scoring RMSD duplicates. 
enrichment_lower <- enricher(
  gene = lower_id,
  # Whether a term is enriched or not is decided through a hypergeometric test against the background distribution. Here, the background distribution is all pooled unique entries from the dataframe. 
  universe = bg,
  # Specify TERM2GENE to allow annotation of the STRING-ID's
  TERM2GENE = annot) |> 
  # Visualize using a dataframe
  as.data.frame() |> 
  # Only select entries that have an adjusted p-value lower than 5% (confidence threshold)
  filter(p.adjust <0.05)

# Pool all duplicates into one vector and select unique entries to avoid duplicates 
upper_id <- c(upper$string_id_dup1, upper$string_id_dup2) |> unique()

# Perform enrichment analysis on the top 10% highest-scoring RMSD duplicates. 
enrichment_upper <- enricher(
  gene = upper_id,
  # Whether a term is enriched or not is decided through a hypergeometric test against the background distribution. Here, the background distribution is all pooled unique entries from the dataframe. 
  universe = bg,
  # Specify TERM2GENE to allow annotation of the STRING-ID's
  TERM2GENE = annot) |> 
  # Visualize using a dataframe
  as.data.frame() |> 
  # Only select entries that have an adjusted p-value lower than 5% (confidence threshold)
  filter(p.adjust <0.05)
```

# B factor computation
Based on visual analysis of two compared structures, there was a hypothesis that the confidence of Alphafold may confound the degree of divergence (based on RMSD) between structures. To test this hypothesis, the B-factor was previously extracted and stored in a dataframe, along with the sequence similarity and RMSD, when running the `get_values` function. 

## Alternative average B-factor computation
The B-factor computed during the `get_values` function is the average B-factor per pair of homologous, at the positions where both structures were aligned and therefore compared. In order to confirm that this is indeed the B-factor, the confidence score is also computed via an alternative way. Here, the confidence score is extracted, per structure, from the Alphafold server via API. Considering the computation of the average B-factor at positions where only both of the sequences is aligned is a rather intensive process, it would also be interesting to check whether this type of computation makes a difference relative to simply extracting the confidence score from the Alphafold server. 

In this step, a function called `get_average_plddt` is set up. This function extracts the confidence score from the Alphafold website per protein ID. 
```{r}
# Define function to extract average pLDDT score from AlphaFoldDB
get_average_plddt <- function(protein_id) {
    # Per protein_id, retrieve the URL to the page of the protein
    url <- paste0('https://alphafold.ebi.ac.uk/api/prediction/', protein_id)
    # Using the httr-library, extract the contents from the URL and retrieve the confidence score (globalMetricValue)
    score <- httr::content(httr::GET(url))[[1]]$globalMetricValue
    return(score)
}
```

With the function specified, the confidence score is extracted for every protein ID considered in previous datasets.
```{r}
# Get all protein entries within the dataset
proteins <- unique(c(data_ordered$dup1_uniprot, data_ordered$dup2_uniprot))

# Per protein entry, get the confidence from the Alphafold website 
plddt_df <- lapply(proteins, function(x) {
  score <- tryCatch(
    # Execute expression; in this case, the function get_average_plddt, which is specified earlier
    expr = {get_average_plddt(x)},
    # If an error occurs, return 'NA'
    error = function(e){return(NA)}
  )
  # Create a dataframe with the protein entry and their respective confidence score, retrieved from the Alphafold website
  df <- data.frame(id = x, score = score)
}) |>
  # Combine the set of dataframes to form a bigger dataframe 
  bind_rows()
```

## Visual analysis
```{r}
# Specify RMSD grouped by duplication type, color based on duplication type
plot1 <- ggplot(b_factor, aes(x = mean_value, y = rmsd)) +
  # Specify violin plot, change outline's color and size
  geom_point(size = 2, shape = 1, color = "#0D5AC6") +
  # Apply a trendline throughout the plot in order to visualize the general behavior of the datapoints
  geom_smooth(method = lm, se = TRUE, alpha = 0.75, color = "black") +
  # Specify geom_point to have empty dots 
  scale_shape(solid = FALSE) +
  # Specify x- and y-titles 
  labs(x = "Average B-factor", y = "RMSD (Å)") +
  # Customize theme
  theme_minimal() +
  theme(
    # Hide legend
    legend.position = "none",
    # Specify background of panel and provide black outline
    panel.background = element_rect(colour = 'black', size=0.8),
    # Remove grid
    panel.grid = element_blank(),
    # Change size on x- and y-axes respectively
    axis.text.x = element_text(color = "black", size=12),
    axis.text.y = element_text(color = "black", size=12),
    # Specify axes lines' color and size 
    axis.line = element_line(color = "black", size=0.8),
    # Add ticks along the labels on the x- and y-axes
    axis.ticks.x = element_line(color = "black", size=0.8),
    axis.ticks.y = element_line(color = "black", size=0.8),
    # Specify tick length
    axis.ticks.length = unit(0.15, "cm"),
    # Change size of plot subtitles 
    strip.text = element_text(size = 10))

# Specify RMSD grouped by duplication type, color based on duplication type
plot2 <- ggplot(b_factor, aes(x = bavg, y = rmsd)) +
  # Specify violin plot, change outline's color and size
  geom_point(size = 2, shape = 1, color = "#0D5AC6") +
  # Apply a trendline throughout the plot in order to visualize the general behavior of the datapoints
  geom_smooth(method = lm, se = TRUE, alpha = 0.75, color = "black") +
  # Specify geom_point to have empty dots 
  scale_shape(solid = FALSE) +
  # Specify x- and y-titles 
  labs(x = "Average B-factor", y = "RMSD (Å)") +
  # Customize theme
  theme_minimal() +
  theme(
    # Hide legend
    legend.position = "none",
    # Specify background of panel and provide black outline
    panel.background = element_rect(colour = 'black', size=0.8),
    # Remove grid
    panel.grid = element_blank(),
    # Change size on x- and y-axes respectively
    axis.text.x = element_text(color = "black", size=12),
    axis.text.y = element_text(color = "black", size=12),
    # Specify axes lines' color and size 
    axis.line = element_line(color = "black", size=0.8),
    # Add ticks along the labels on the x- and y-axes
    axis.ticks.x = element_line(color = "black", size=0.8),
    axis.ticks.y = element_line(color = "black", size=0.8),
    # Specify tick length
    axis.ticks.length = unit(0.15, "cm"),
    # Change size of plot subtitles 
    strip.text = element_text(size = 10))

figure4 <- grid.arrange(plot1, plot2, ncol=2)
```

## Statistical analysis
```{r}
fit <- lm(data = b_factor, rmsd ~ -bavg)
summary(fit)
report(fit)
```

```{r}
# 
b_factor <- b_factor %>%
  # Get the B-factor for dup1_uniprot entries from the previously gathered confidence scores from the Alphafold server
  left_join(plddt_df, by = c("dup1_uniprot" = "id")) %>%
  # Rename the column to avoid collision with the computing of the B-factor for the dup2_uniprot entries 
  rename(dup1_score = score) %>%
  # Get the B-factor for dup2_uniprot entries from the previously gathered confidence scores from the Alphafold server
  left_join(plddt_df, by = c("dup2_uniprot" = "id")) %>%
  # Rename the column
  rename(dup2_score = score) %>%
  # Compute the average B-factor score per pair of homologous proteins 
  mutate(mean_value = (dup1_score + dup2_score) / 2)


b_factor <- b_factor %>%
  mutate(diff = (bavg - mean_value))


b_factor <- merge(
  b_factor,
  data_idx[, c("dup1_uniprot", "dup2_uniprot", "idx")],   # only keep the needed column(s)
  by = c("dup1_uniprot", "dup2_uniprot"),               # match using two columns
  all.x = TRUE)
```

```{r}
# Specify RMSD grouped by duplication type, color based on duplication type
figure5 <- ggplot(b_factor, aes(x = idx, y = diff)) +
  # Specify violin plot, change outline's color and size
  geom_point(size = 2, shape = 1, color = "#0D5AC6") +
  # Apply a trendline throughout the plot in order to visualize the general behavior of the datapoints
  geom_smooth(method = lm, se = TRUE, alpha = 0.75, color = "black") +
  # Specify geom_point to have empty dots 
  scale_shape(solid = FALSE) +
  # Specify x- and y-titles 
  labs(x = "Index", y = "Difference Average B-factor") +
  # Customize theme
  theme_minimal() +
  theme(
    # Hide legend
    legend.position = "none",
    # Specify background of panel and provide black outline
    panel.background = element_rect(colour = 'black', size=0.8),
    # Remove grid
    panel.grid = element_blank(),
    # Change size on x- and y-axes respectively
    axis.text.x = element_text(color = "black", size=12),
    axis.text.y = element_text(color = "black", size=12),
    # Specify axes lines' color and size 
    axis.line = element_line(color = "black", size=0.8),
    # Add ticks along the labels on the x- and y-axes
    axis.ticks.x = element_line(color = "black", size=0.8),
    axis.ticks.y = element_line(color = "black", size=0.8),
    # Specify tick length
    axis.ticks.length = unit(0.15, "cm"),
    # Change size of plot subtitles 
    strip.text = element_text(size = 10))
figure5
```

# Report sessioninfo
```{r}
session_info(
  pkgs = c("attached", "installed")[1],
  include_base = FALSE,
  info = c("auto"),
  to_file = FALSE
)
```

